---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MMSpower

<!-- badges: start -->
[![R-CMD-check](https://github.com/mrc-ide/MMSpower/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mrc-ide/MMSpower/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/mrc-ide/MMSpower/graph/badge.svg)](https://app.codecov.io/gh/mrc-ide/MMSpower)
<!-- badges: end -->

The goal of MMSpower is to ...

## !!! UNRESOLVED - `design_detection()` does not use specificity

**This must be addressed before `design_detection()` is treated as
validated.** The function currently computes the per-sample detection
probability as

```
q = prevalence * sensitivity
```

and **ignores `specificity` entirely** (it is validated and echoed back,
but does not enter the formula). This follows the standard surveillance /
"freedom from disease" basis (Cannon & Roe 1982; Cannon 2001), where
detection means catching a *true* occurrence and false positives are
handled separately.

Two decisions are still open and need the team:

1. Should "detection" mean **"detect a true occurrence"** (current,
   `q = p * Se`) or **"observe any positive result"** (the
   apparent-prevalence basis `q = p*Se + (1-p)*(1-Sp)` used by every other
   function in this package via `.apparent_prev()`)?
2. If false positives are in scope, how should a "detection" that is
   actually a false positive be reported or discounted?

Until this is settled, treat any `design_detection()` result with
`specificity < 1` as provisional. See the "Known limitation" section in
`?design_detection`.

## Installation

You can install the development version of MMSpower from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("mrc-ide/MMSpower")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(MMSpower)
## basic example code
```

