---
title: "Getting started with MMSpower"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with MMSpower}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(MMSpower)
```

## The idea

MMSpower supports the two halves of a malaria molecular surveillance
study:

1. **Design** -- before collecting data, work out how many samples, sites,
   or survey rounds you need.
2. **Analysis** -- once the data are in, produce an estimate, test, or
   confidence interval.

Each study question has a function for each half:

| Question | Design | Analysis |
|---|---|---|
| Prevalence and its precision | `design_precision()` | `estimate_prevalence()` |
| Above/below a threshold | `design_threshold()` | `test_threshold()` |
| Difference between two groups | `design_difference()` | `test_difference()` |
| Trend over time | `design_trend()` | `test_trend()` |
| Detect a rare variant | `design_detection()` | -- |
| Study cost | `estimate_cost()` | -- |

All of them take the same three adjustments, so a design and its analysis
stay consistent:

* `sensitivity`, `specificity` -- Rogan-Gladen correction for an imperfect
  assay;
* `icc` (or `n_per_site` / `n_sites`) -- a Kish design effect for
  clustered sampling;
* `fpc_N` -- an optional finite-population correction.

## A worked example: prevalence

Suppose we plan a survey to estimate the prevalence of a resistance marker
to within 5 percentage points, expecting roughly 15% prevalence, sampling
30 people at each of several health facilities with a modest amount of
between-facility correlation.

```{r design, eval = FALSE}
# design_precision() lives on its own feature branch for now
design_precision(
  prevalence = 0.15,
  moe        = 0.05,
  n_per_site = 30,
  icc        = 0.05
)
```

After running the survey we have per-site positive counts and totals, and
we know the assay is not perfect. `estimate_prevalence()` corrects for the
assay, estimates the intra-cluster correlation from the spread between
sites, and reports both the raw and effective sample sizes.

```{r analysis}
res <- estimate_prevalence(
  x = c(2, 14, 4, 18, 6),
  n = c(60, 60, 55, 65, 50),
  sensitivity = 0.90,
  specificity = 0.98
)

res[c("prevalence", "ci_lower", "ci_upper", "deff", "icc_used",
      "n_total", "n_eff")]
```

Here the five sites vary more than sampling alone explains
(`icc_used` `r round(res$icc_used, 3)`), so the design effect
(`r round(res$deff, 1)`) inflates and the `r res$n_total` samples are
worth only about `r round(res$n_eff)` independent observations. A design
that assumed independent sampling would have been badly under-powered.

## Where the methods come from

Every function documents its equations and cites the MMS-SD Study Design
Workshop module they come from, plus the standard references for the parts
that go beyond the workshop. See `?estimate_prevalence` (the "Equations
and sources" section) and:

* `vignette("methods")` -- the full derivation for the prevalence /
  precision pair;
* `vignette("methods-and-sources")` -- the cross-function equation-to-
  lecture index.
