---
title: "Methods: prevalence estimation and its precision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods: prevalence estimation and its precision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(MMSpower)
```

## Scope

This vignette documents every equation used by `estimate_prevalence()`, the
analysis function that turns observed test-positive counts into a prevalence
point estimate and confidence interval (CI). It covers *why* each adjustment
is applied, not only the formula, and ends with a worked example that shows
the intermediate quantities.

`estimate_prevalence()` is the analysis counterpart of the design function
`design_precision()`: the former reports the precision a completed survey
achieved, the latter sizes a survey to hit a target precision. They share the
same variance model, so the two round-trip (see
[Relationship to `design_precision()`](#relationship-to-design_precision)).

## Notation

| Symbol | Meaning | In the return value |
|---|---|---|
| $x_i,\ n_i$ | positive count and total in cluster $i$ | inputs `x`, `n` |
| $\hat{p}$ | apparent (observed-test) prevalence | — |
| $p$ | true prevalence | `prevalence` |
| $Se,\ Sp$ | diagnostic sensitivity, specificity | `sensitivity`, `specificity` |
| $\bar{n}$ | mean cluster size, $\frac{1}{k}\sum_i n_i$ | — |
| $\rho$ | intra-cluster correlation (ICC) | `icc_used` |
| $D_{eff}$ | Kish design effect | `deff` |
| $n_{total}$ | $\sum_i n_i$ | `n_total` |
| $n_{eff}$ | effective (independent-equivalent) sample size | `n_eff` |
| $N$ | population size (headcount) for the finite-population correction | input `fpc_N` |
| $f$ | finite-population variance factor, $(N - n_{total})/(N - 1)$ | — |
| $n_{eff,adj}$ | effective sample size after the FPC | — |
| $z$ | $\Phi^{-1}(1 - \alpha/2)$, with $\alpha = 1 -$ `conf_level` | — |

## The pipeline

The interval is assembled in six stages. Each stage feeds the next:

1. apparent prevalence $\hat{p}$
2. design effect $D_{eff}$ (from a supplied or estimated ICC)
3. effective sample size $n_{eff} = n_{total} / D_{eff}$
4. finite-population correction $\rightarrow n_{eff,adj}$
5. confidence interval on $\hat{p}$ by the chosen `method`
6. Rogan-Gladen correction $\hat{p} \rightarrow p$, applied to the point
   estimate and both endpoints

### 1. Apparent prevalence

$$
\hat{p} \;=\; \frac{\sum_i x_i}{\sum_i n_i}
$$

The pooled proportion of test positives. Every CI method operates on
$\hat{p}$; the correction for an imperfect test (stage 6) is deferred to the
very end so that it is applied *identically* to the point estimate and to
both interval endpoints — an affine map of an interval is still an interval,
so order matters only for clarity, not correctness.

### 2. Design effect (Kish)

Two people sampled from the same village give more similar results than two
people sampled from the whole country: within-cluster observations are
positively correlated. A clustered sample of $n_{total}$ observations
therefore carries the information of *fewer* than $n_{total}$ independent
ones. The Kish (1965) design effect quantifies the inflation of the variance:

$$
D_{eff} \;=\; 1 + (\bar{n} - 1)\,\rho
$$

where $\rho \in [0, 1]$ is the intra-cluster correlation. $\rho = 0$ (or a
single observation per cluster, $\bar{n} = 1$) gives $D_{eff} = 1$ and the
simple-random-sample (SRS) case. Taking $\rho = 0$ *by default* would
understate uncertainty for any genuinely clustered survey, so the function
does not: it estimates $\rho$ unless you supply one.

#### 2a. ICC supplied (`icc` set)

$\rho$ is used directly in the formula above. Use this when you have an
external estimate of the ICC (e.g. from a previous survey) or want to run a
sensitivity analysis across plausible values.

#### 2b. ICC estimated (`icc = NULL`, the default)

With $k \ge 2$ clusters that are not all of size 1, the design effect is
estimated as the ratio of the **observed** between-cluster variance of the
site proportions $\hat{p}_i = x_i / n_i$ to the variance **expected under
SRS**:

$$
\widehat{D_{eff}}
  \;=\;
  \frac{\mathrm{Var}(\hat{p}_i)}
       {\dfrac{1}{k}\sum_i \dfrac{\hat{p}\,(1 - \hat{p})}{n_i}}
$$

The numerator is the empirical scatter of the cluster-level proportions; the
denominator is what that scatter *would* be if the clusters differed only by
binomial sampling noise. Their ratio is, by definition, the design effect.
It is floored at 1 (clustering cannot buy you information). The implied ICC
is recovered by inverting the Kish formula,

$$
\hat{\rho} \;=\; \frac{\widehat{D_{eff}} - 1}{\bar{n} - 1},
$$

and clamped to $[0, 1]$. $D_{eff}$ is then **recomputed** from the clamped
$\hat{\rho}$ so that the returned `deff` and `icc_used` are always mutually
consistent (they satisfy the Kish identity exactly).

When there is a single cluster, or every cluster has size 1, the Kish
denominator $\bar{n} - 1$ is 0 and the ICC is not identifiable; the function
falls back to $D_{eff} = 1$.

> **Caveat.** The variance-ratio estimator is noisy with few clusters. With,
> say, 3–5 sites it can land at the $[0, 1]$ clamp or swing between runs.
> If you have a trustworthy external ICC, supplying it is more stable.

### 3. Effective sample size

$$
n_{eff} \;=\; \frac{n_{total}}{D_{eff}}
$$

the number of *independent* observations that would carry the same
information as the clustered sample. This is the quantity every CI method
actually uses in place of a raw count. Returned as `n_eff`.

### 4. Finite-population correction

The usual binomial variance $\hat{p}(1-\hat{p})/n$ assumes sampling with
replacement from an infinite population. When you sample a non-trivial
fraction of a **finite** population of known size $N$ (`fpc_N`), sampling
*without* replacement leaves less residual uncertainty — in the limit where
you sample everyone, there is none. The variance is multiplied by

$$
f \;=\; \frac{N - n_{total}}{N - 1} \;\in\; [0, 1).
$$

The FPC uses the **collected count** $n_{total}$ — the real sampling
fraction $n_{total}/N$ — not the design-effect-adjusted $n_{eff}$. The
design effect and the FPC are independent: $D_{eff}$ measures how
*inefficient* the design was, the FPC measures how *much of the
population* you observed. (`fpc_N` is a headcount, so the population is
treated as finite in individuals; a study that sampled nearly all *sites*
would also shrink the between-cluster variance, which this single-FPC
shortcut does not separately model.) `design_precision()` and
`design_threshold()` apply the FPC the same way.

Instead of carrying $f$ through three different variance expressions, it is
folded once into an adjusted sample size:

$$
n_{eff,adj} \;=\; \frac{n_{eff}}{f}
$$

because dividing the variance's denominator by $f$ is the same as
multiplying the variance by $f$:
$\dfrac{\hat{p}(1-\hat{p})}{n_{eff}} \cdot f
 = \dfrac{\hat{p}(1-\hat{p})}{n_{eff}/f}
 = \dfrac{\hat{p}(1-\hat{p})}{n_{eff,adj}}$.
All three CI methods then use $n_{eff,adj}$ and nothing else changes. With no
FPC (`fpc_N = NULL`), $f = 1$ and $n_{eff,adj} = n_{eff}$.

As $n_{total} \to N$, $f \to 0$ and $n_{eff,adj} \to \infty$: the sampling
variance vanishes. The function stops with an informative error before that
point (you have effectively surveyed the whole population; no CI is needed).

### 5. Confidence interval on apparent prevalence

Let $z = \Phi^{-1}(1 - \alpha/2)$ with $\alpha = 1 -$ `conf_level`, and
define the effective positive count $x_{eff} = \hat{p}\,n_{eff,adj}$ (a
continuous quantity — it need not be an integer, which is what lets the
clustering and FPC adjustments flow through the exact method below).

#### `"wald"` (default)

$$
\hat{p} \;\pm\; z\,\sqrt{\frac{\hat{p}\,(1 - \hat{p})}{n_{eff,adj}}}
$$

clamped to $[0, 1]$. The textbook normal approximation. Fast, symmetric, and
what most readers expect. Its weaknesses: it **under-covers** (true coverage
< nominal) for small $n_{eff,adj}$ or prevalence near 0 or 1, and it
degenerates to a zero-width interval when $\hat{p} = 0$ or $1$.

#### `"clopper-pearson"`

The exact binomial interval, obtained through the beta-quantile identity
(the binomial CDF inverts to a beta quantile):

$$
L = B^{-1}\!\left(\tfrac{\alpha}{2};\; x_{eff},\; n_{eff,adj} - x_{eff} + 1\right)
$$
$$
U = B^{-1}\!\left(1 - \tfrac{\alpha}{2};\; x_{eff} + 1,\; n_{eff,adj} - x_{eff}\right)
$$

with $L = 0$ when $\hat{p} = 0$ and $U = 1$ when $\hat{p} = 1$, then clamped
to $[0, 1]$. "Exact" means coverage is **guaranteed** to be at least
nominal for every true $p$ — never below 95% for a 95% interval. The price
is conservatism: the interval is wider than it strictly needs to be. Use it
for small samples or extreme prevalences, where Wald's under-coverage
matters and a guarantee is worth the width.

#### `"agresti-coull"`

Add $z^2$ pseudo-observations ($\approx 4$ at 95%, split evenly between
"positive" and "negative"), then take a Wald interval on the resulting
adjusted proportion:

$$
\tilde{n} = n_{eff,adj} + z^2,
\qquad
\tilde{p} = \frac{x_{eff} + z^2/2}{\tilde{n}}
$$
$$
\tilde{p} \;\pm\; z\,\sqrt{\frac{\tilde{p}\,(1 - \tilde{p})}{\tilde{n}}}
$$

clamped to $[0, 1]$. The pseudo-count pulls $\tilde{p}$ toward $1/2$, which
fixes most of Wald's boundary misbehaviour while keeping a simple
closed form. Coverage is close to nominal for moderate $n$ without
Clopper-Pearson's systematic over-width — a sensible middle-ground default.

#### Choosing a method

| Situation | Method |
|---|---|
| Moderate/large $n$, prevalence away from 0/1, want the familiar interval | `"wald"` |
| Need a coverage **guarantee**; small $n$ or very low/high prevalence | `"clopper-pearson"` |
| Small–moderate $n$, want good coverage without over-width | `"agresti-coull"` |

### 6. Rogan-Gladen correction (imperfect test)

A test with sensitivity $Se < 1$ misses some true positives; specificity
$Sp < 1$ adds some false positives. The apparent prevalence is a known
linear function of the true prevalence:

$$
p_{app} \;=\; p\,Se \;+\; (1 - p)\,(1 - Sp).
$$

Solving for $p$ gives the Rogan-Gladen (1978) estimator:

$$
p \;=\; \frac{p_{app} - (1 - Sp)}{Se + Sp - 1}
$$

This affine map is applied to $\hat{p}$ **and** to both CI endpoints from
stage 5, and each result is clamped to $[0, 1]$. When $Se = Sp = 1$ it is the
identity and nothing happens.

The denominator $Se + Sp - 1$ is the Youden index; it must be positive (the
test must do better than a coin flip) or true prevalence is not identifiable
from apparent prevalence — the function errors. As $Se + Sp - 1 \to 0$ the
correction divides by a vanishing number, so both the point estimate and the
interval blow up; the function warns when it drops below 0.1, where results
are numerically unreliable.

> **Note on the variance.** `estimate_prevalence()` corrects the *endpoints*
> by the affine map, which widens the interval by the factor
> $1/(Se + Sp - 1)$ — the correct first-order (delta-method) behaviour.
> `design_precision()` uses the same factor squared in the variance when
> sizing a study.

### Margin of error

From the corrected point estimate `prevalence` ($p$) and endpoints
`ci_lower`, `ci_upper`:

$$
\texttt{moe} = \frac{\texttt{ci\_upper} - \texttt{ci\_lower}}{2},
\qquad
\texttt{moe\_lower} = p - \texttt{ci\_lower},
\qquad
\texttt{moe\_upper} = \texttt{ci\_upper} - p
$$

For `"wald"` the interval is symmetric about $p$ and all three are equal —
`moe` alone describes the precision. For `"clopper-pearson"` and
`"agresti-coull"` the interval is **asymmetric**: `moe` is only the *average*
of the two half-widths and describes neither side, so report `moe_lower` and
`moe_upper` together. The function emits a `message()` when they differ
noticeably.

## Worked example

Eight clustered sites, an imperfect test ($Se = 0.90$, $Sp = 0.95$), 95%
Clopper-Pearson interval, ICC estimated from the data:

```{r worked}
x <- c(2, 5, 1, 18, 22, 14, 9, 6)
n <- c(60, 80, 70, 100, 40, 60, 50, 90)

res <- estimate_prevalence(
  x = x, n = n,
  sensitivity = 0.90,
  specificity = 0.95,
  method      = "clopper-pearson"
)

str(res)
```

Reproducing the intermediate quantities by hand:

```{r worked-check}
p_hat   <- sum(x) / sum(n)                      # 1. apparent prevalence
n_bar   <- mean(n)
p_i     <- x / n
var_obs <- var(p_i)
var_srs <- mean(p_hat * (1 - p_hat) / n)
deff    <- max(var_obs / var_srs, 1)            # 2. design effect (estimated)
icc     <- min(max((deff - 1) / (n_bar - 1), 0), 1)
deff    <- 1 + (n_bar - 1) * icc                # recomputed for consistency
n_eff   <- sum(n) / deff                        # 3. effective sample size

c(p_hat = p_hat, deff = deff, icc = icc, n_eff = n_eff)

# matches the returned values:
c(deff = res$deff, icc_used = res$icc_used, n_eff = res$n_eff)
```

```{r worked-rg}
# 6. Rogan-Gladen: apparent -> true, applied to the point estimate
Se <- 0.90; Sp <- 0.95
(p_hat - (1 - Sp)) / (Se + Sp - 1)   # == res$prevalence (up to clamping)
res$prevalence
```

Because the method is Clopper-Pearson, the interval is asymmetric — note
`moe_lower` and `moe_upper` differ, and `moe` sits between them:

```{r worked-moe}
c(moe_lower = res$moe_lower, moe = res$moe, moe_upper = res$moe_upper)
```

## Relationship to `design_precision()`

`design_precision()` inverts the Wald variance to size a study for a target
margin of error `moe`:

$$
n \;=\; \frac{z^2\, p_{app}(1 - p_{app})\, D_{eff}}
              {\texttt{moe}^2\,(Se + Sp - 1)^2}
$$

Sizing with `design_precision(prevalence = p, moe = m, ...)` and then
analysing a dataset of that size with `estimate_prevalence(..., method =
"wald")` returns `moe` $\approx m$ — the same $z$, the same $D_{eff}$
model, the same Rogan-Gladen factor. The two functions are the design-time
and analysis-time views of one variance model.

The round-trip is exact only for `method = "wald"`. Clopper-Pearson and
Agresti-Coull produce asymmetric intervals whose average half-width is close
to, but not identical to, the Wald `moe` that `design_precision()` targets.

## References

- Kish, L. (1965). *Survey Sampling*. Wiley.
- Clopper, C. J. & Pearson, E. S. (1934). The use of confidence or fiducial
  limits illustrated in the case of the binomial. *Biometrika*, 26(4),
  404–413. <https://doi.org/10.1093/biomet/26.4.404>
- Agresti, A. & Coull, B. A. (1998). Approximate is better than "exact" for
  interval estimation of binomial proportions. *The American Statistician*,
  52(2), 119–126. <https://doi.org/10.1080/00031305.1998.10480550>
- Rogan, W. J. & Gladen, B. (1978). Estimating prevalence from the results
  of a screening test. *American Journal of Epidemiology*, 107(1), 71–76.
  <https://doi.org/10.1093/oxfordjournals.aje.a112510>
